---
title: "Third Assignment"
author: "Emilia Sicari & Rafael Lopez V."
date: "March 25, 2016"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
  html_document:
    toc: yes
    toc_depth: '2'
subtitle: Christopher Gandrud, Hertie School of Governance, Spring 2016
bibliography: packages.bib
---

```{r setup, include=FALSE, error=FALSE, message=FALSE, results = 'asis'}
knitr::opts_chunk$set(echo = TRUE) 

#setting working directory
getwd()
try(setwd("/Users/emiliasicari/Desktop/Third_assignment"), silent = TRUE)
try(setwd("/Users/rafalopezv/Dropbox/R/Third-assignment/"), silent = TRUE)

#loading packages or installing them if not already available
library(devtools)
library(Quandl)
library(data.table)
library(rio)
library(ggplot2)
library(stargazer)
library(pander)
library(knitr)
library(scales)

# creating an object of the packages for citation purposes
packages <- c('devtools', 'repmis', 'Quandl', 'data.table', 'rio', 'ggplot2', 'stargazer', 'pander', 'Scale')
repmis::LoadandCite(packages, file = 'packages.bib')

# importing data on Singapore's economic growth, selecting only the time span of interest
URL <- "https://www.quandl.com/api/v3/datasets/ODA/SGP_NGDPPC.csv"
gdp.pc <- repmis::source_data(URL, header = TRUE, nrows=21, skip = 5)
#adding a new column, and inserting content 
gdp.pc["MY_NEW_COLUMN"] <- 2015:1995
#erasing a column
gdp.pc$`2016-12-31`<- NULL
#changing the names of the columns
names(gdp.pc) <- c("gdp per capita", "date")
#changing order of columns
gdp.pc[,c(2,1)]
#checking the class of the varibles
sapply(gdp.pc, class)

#importing data on Singapore's bottom 90% average income, selecting only the time span of interest
bottom <- read.csv("ObservationData_zqpldgb.csv", header = TRUE, skip = 20)
#erasing columns
bottom$Singapore <- NULL
bottom$Bottom.90..average.income <- NULL
bottom$Real.2010.Singapore.Dollars <- NULL
bottom$X1.1.1994.12.00.00.AM <- NULL
#adding a new column and inserting content
bottom["MY_NEW_COLUMN"] <- 1995:2009
#rename columns
names(bottom) <- c("bottom", "date")
#reorder columns
bottom[,c(2,1)]
#checking the class of the varibles
sapply(bottom, class)
#forecasrting the bottom 90% average income for the missing years (2010-2015) and creating a dataframe with the forecasted values
Forecast <- lm(bottom  ~ date, data = bottom)
116.7*2010+(-204924.2) #forcasted bottom 90% average income for 2010
116.7*2011+(-204924.2) #forcasted bottom 90% average income for 2011
116.7*2012+(-204924.2) #forcasted bottom 90% average income for 2012
116.7*2013+(-204924.2) #forcasted bottom 90% average income for 2013
116.7*2014+(-204924.2) #forcasted bottom 90% average income for 2014
116.7*2015+(-204924.2) #forcasted bottom 90% average income for 2015
new.bottom <- data.frame(date = c(2010, 2011, 2012, 2013, 2014, 2015), bottom = c(29642.8, 29759.5, 29876.2, 29992.9, 30109.6, 30226.3)) #creating a new dataframe with the forecastedvalues
new.bottom #visualising the new dataframe
#binding the new.bottom dataframe with the original bottom dataframe
bottom.complete <- rbind(bottom, new.bottom)

#importing data on Singapore's top 10% average income, selecting only the time span of interest
top <-read.csv("ObservationData_rblkfmd.csv", header = TRUE, skip = 20)
#erasing columns
top$Singapore <- NULL
top$Top.10..average.income <- NULL
top$Real.2010.Singapore.Dollars <- NULL
top$X1.1.1994.12.00.00.AM <- NULL
#adding a new column and inserting content
top["MY_NEW_COLUMN"] <- 1995:2009
#rename columns
names(top) <- c("top", "date")
#reorder columns
top[,c(2,1)]
#checking the class of the varibles
sapply(top, class)
#forecasrting the top 10% average income for the missing years (2010-2015) and creating a dataframe with the forecasted values
Forecast2 <- lm(top ~ date, data = top)
Forecast2
6312*2010-12476918  #forcasted top 10% average income for 2010
6312*2011-12476918  #forcasted top 10% average income for 2011
6312*2012-12476918  #forcasted top 10% average income for 2012
6312*2013-12476918  #forcasted top 10% average income for 2013
6312*2014-12476918  #forcasted top 10% average income for 2014
6312*2015-12476918  #forcasted top 10% average income for 2015
new.top <- data.frame(date = c(2010, 2011, 2012, 2013, 2014, 2015), top = c(210202, 216514, 222826, 229138, 235450, 241762)) #creating a new dataframe with the forecastedvalues
new.top #visualising the new dataframe
#binding the new.top dataframe with the original top dataframe
top.complete <- rbind(top, new.top)

#creating a new dataset showing the variation between the top 10% and bottom 90% average income over time, to measure inequality between 1995 and 2015
inequality.complete <- data.frame(date = c(1995:2015), inequality = c(top.complete$top / bottom.complete$bottom))
                 
#Downloading data set for public transport utilisation (from 1995 to 2015)
a <- "https://data.gov.sg/dataset/7f0661dc-6e4f-460f-8e8c-3514c9a8cb6b/resource/552b8662-3cbc-48c0-9fbb-abdc07fb377a/download/public-transport-utilisation-average-public-transport-ridership.csv"
trans.ut <- repmis::source_data(a)

#Changing the name of the colums 
names(trans.ut) <- c("date","type","average.1000.day")

#Checking the class of the data
sapply(trans.ut, class)

# Subsetting the "trans.ut file" to make it compatible with the other data and foreseeing the merging process 
trans.ut.mrt <- data.frame(subset(trans.ut, type == "MRT"))
trans.ut.bus <- data.frame(subset(trans.ut, type == "Bus"))
trans.ut.lrt <- data.frame(subset(trans.ut, type == "LRT"))
trans.ut.taxi <- data.frame(subset(trans.ut, type == "Taxi"))

# Completing the missing data for the trans.ut.lrt data frame
trans.ut.lrt1 <- data.frame(date = c(1995,1996,1997,1998), type = c("LRT", "LRT","LRT", "LRT"), average.1000.day = c(0,0,0,0))

# Binding data sets: trans.ut.lrt and trans.ut.lrt1  
trans.ut.lrt2 <- rbind(trans.ut.lrt, trans.ut.lrt1)
trans.ut.lrt2
#Downloading data for car population (from 1965 to 2015)
b <- "https://data.gov.sg/dataset/e2673131-f79f-43e3-9579-c17cc25e1735/resource/31ca0cee-6d9e-453a-8b4f-376d37713a10/download/motor-vehicle-population-by-type-of-vehicle-end-of-period-annual.csv"
car.pop <- repmis::source_data(b)

#Changing the name of the colums
names(car.pop) <- c("date", "type", "number")

# Eliminating the years that are not compatible with the other data sets
car.pop1 <- data.frame(subset(car.pop, date > "1994"))
car.pop1 <- data.frame(subset(car.pop1, date < "2015"))

#Checking the class of the data
sapply(car.pop1, class)

#Transform the data of one of the columns that have been imported as characters
car.pop1 <- transform(car.pop1, number = as.numeric(number))

# Subsetting the "car.pop1 file" to make it compatible with the other data and foreseeing the merging process 
car.pop1.cars <- data.frame(subset(car.pop1, type == "Cars"))
car.pop1.rentalcars <- data.frame(subset(car.pop1, type == "Rental Cars"))
car.pop1.taxis <- data.frame(subset(car.pop1, type == "Taxis"))
car.pop1.buses <- data.frame(subset(car.pop1, type == "Buses"))
car.pop1.motorbikes <- data.frame(subset(car.pop1, type == "Motorcycles & Scooters"))
car.pop1.other <- data.frame(subset(car.pop1, type == "Goods & Other Vehicles"))

#changing name of the column "year" into "date"
names(car.pop1.rentalcars) <- c("date", "type", "number")
names(car.pop1.cars) <- c("date", "type", "number")
names(car.pop1.taxis) <- c("date", "type", "number")
names(car.pop1.buses) <- c("date", "type", "number")
names(car.pop1.motorbikes) <- c("date", "type", "number")
names(car.pop1.other) <- c("date", "type", "number")

#merging inequality and gdp per capita dataframes
gdp.ineq <- merge(gdp.pc, inequality.complete, by ='date')
#eliminating the extra year (2015) in order to proceed with merging with the other data frames
gdp.ineq1 <- data.frame(subset(gdp.ineq, date<"2015"))
bottom.complete <- data.frame(subset(bottom.complete, date<"2015"))
top.complete <- data.frame(subset(top.complete, date<"2015"))
#merging gdp.ineq dataframe with all the other dataframes, and cleaning meanwhile

#merging with cars and cleaning
dataframe1 <- merge (gdp.ineq1, car.pop1.cars, by ='date')
dataframe1$type <- NULL
names(dataframe1) <- c("date", "gdp.per.capita", "inequality", "cars")

#merging with rentalcars and cleaning
dataframe2 <- merge (dataframe1, car.pop1.rentalcars, by ='date')
dataframe2$type <- NULL
names(dataframe2) <- c("date", "gdp.per.capita", "inequality", "cars", "rentalcars")

#merging with taxis and cleaning
dataframe3 <- merge (dataframe2,car.pop1.taxis, by ='date')
dataframe3$type <- NULL
names(dataframe3) <- c("date", "gdp.per.capita", "inequality", "cars", "rentalcars", "taxis")

#merging with buses and cleaning
dataframe4 <- merge (dataframe3, car.pop1.buses, by ='date')
dataframe4$type <- NULL
names(dataframe4) <- c("date", "gdp.per.capita", "inequality", "cars", "rentalcars", "taxis", "buses")

#merging with motorbikes and cleaning
dataframe5 <- merge (dataframe4, car.pop1.motorbikes, by ='date')
dataframe5$type <- NULL
names(dataframe5) <- c("date", "gdp.per.capita", "inequality", "cars", "rentalcars", "taxis", "buses", "motorbikes")

#creating the almost final, complete dataframe and cleaning it
almost.complete.dataframe <- merge (dataframe5, car.pop1.other, by ='date')
almost.complete.dataframe$type <- NULL
names(almost.complete.dataframe) <- c("date", "gdp.per.capita", "inequality", "cars", "rentalcars", "taxis", "buses", "motorbikes", "other")

# Eliminating unnecessary colums of the data frames trans.ut.(x) and changing the names

#buses utilization
trans.ut.bus$type <- NULL
names(trans.ut.bus) <- c("date", "bus.u") 

#mrt utilization
trans.ut.mrt$type <- NULL
names(trans.ut.mrt) <- c("date", "mrt.u") 

#lrt utilization
trans.ut.lrt2$type <- NULL
names(trans.ut.lrt2) <- c("date", "lrt.u")

# Creating the very final data frame, inserting the missing data on utilization, as well as bottom 90% and top 10% average income
data.final1 <- merge (almost.complete.dataframe, trans.ut.bus, by ='date')
data.final2 <- merge (data.final1, trans.ut.mrt, by ='date')
data.final3 <- merge (data.final2, top.complete, by='date')
data.final4 <- merge (data.final3, bottom.complete, by='date')
data.final <- merge (data.final4, trans.ut.lrt2, by ='date')

# Moving the columns
data.final[,c(1,2,3,12,13,4,5,6,7,8,9,10,11,14)]
data.final

# Exporting the final data frame as csv file
rio::export(data.final, "final.data.frame.csv", col.names = TRUE)

```

\pagebreak
#Research question and project description
In our collaborative research project, we seeks to provide an answer to the following question: **How the rise in inequality and economic growth influences the purchase of cars as an example of luxury goods, compared to the usage of public transportation systems in Singapore, from 1995 to 2014**. 

Therefore, we collected data on economic growth, income inequality, usage of public transports and purchase of cars covering the time span of 19 years, from 1995 until 2014. As suggested by our research question, economic growth, income inequality are the explanatory variables, while purchase of cars and usage of public transports are the dependent variables. The reason why we chose cars as example of luxury goods showing social status, is that in Singapore purchasing of cars is particularly expensive, due to high taxation and a certificate of car entitlement, whose cost can be higher than 70.000 dollars. For more details about the research proposal and case justification see [ResearchProposal](https://github.com/EmiliaSicari/ResearchProposal).

#Processing data

##Data sources and data gathering
The data that we need for our empirical analysis were retrieved from:

* IMF Cross Country Macroeconomic Statistics open data available on [Quandl](https://www.quandl.com/data/ODA/SGP_NGDPPC) From this source we downloaded data showing the trend in Singopore's GDP per capita meassured in singaporean dollars from 1995 to 2014. The data was provided in csv format.  

* [Knoema](http://knoema.com/). From The World Top Incomes Database - providing access to data on the distribution of top incomes in more than twenty five countries across the globe - we downloaded data on the top 10% average income and bottom 90% average income in Singapore, meassured in singaporean dollars. The reason why we did not gather data from the database [Clio Infra](https://www.quandl.com/data/CLIO/SGP_II-Income-Inequality-Singapore) available on Quandl as we had defined in our [ResearchProposal](https://github.com/EmiliaSicari/ResearchProposal) , is that it did not provide sufficient data for the time span we are considering. Since it was not possible to directly download the data base to R we requested the data, which have been sent via e-mail in csv format. This data is part of the repository.  

* [Singapore's open data portal](https://data.gov.sg). This portal offered two data bases: 
    + [Annual Motor Vehicle Population](https://data.gov.sg/dataset/annual-motor-vehicle-population-by-type-of-vehicles). The data provides the number of cars from 1995 to 2014, including: mortorbikes, rental cars, buses, taxis and other type of vehicles. The last two are considered public cars since in Singapore even the taxis are provided by the state.
    
    + [Public transport utilization](https://data.gov.sg/dataset/public-transport-utilisation-average-public-transport-ridership). This data is expressed as the daily average of thounsand commuters using public trasport by year. It covers the span from 1995 to 2014 and includes the following modes of transportation: MRT (underground), LRT (similar to the Berlin S-Bahn), Taxis (publicly run), Buses.
    
The following table summarizes the variables downloaded.

```{r, echo = FALSE, message=FALSE, warning = FALSE, results='asis'}
# Creating the summary table stated before
Variable <- c("gdp per capita","inequality","anual motor vehicle","public transport utilization")
Description <- c("measured in singaporean dollars at current prices. From 1995 to 2014", "top 10% and bottom 90% singaporean's average income measured in singaporean dollars", "number of: cars, rental cars, buses, taxis, buses, motorbikes", "average commuters using daily: MRT, LRT, Buses, Taxis")  
Time.frame <- c("1980-2014", "1947-2009", "1960-2015", "1995-2014")
data.table <- data.frame(Variable, Description, Time.frame)
pander::pander(data.table, split.cells = 30, split.table = Inf, style = "grid")
```

##Cleaning, processing and merging of datasets

* After importing data we used the "date" variable (year) as a unique identifier for all four data sets, in order to merge them afterwards.

* Since time frames of the data were different, we selected a common span of time: 1995-2014. In one case we had to make a linear regression to forecast mising values to complete the time common frame, and in other we completed the data with we abstent for some years with "0", since we new that the problem was not lack of data but inexistence of that variable in those years.

* Cleaning the data was limited to changing column names, eliminating the unnecesary ones and organizing the data in multile data frames so to merge them more easily  afterwards. In one case we changed the format of the data from characters to integers, due to an incorrect import.

* Data of inequality, namely bottom 90% and top 10% average income, were available only until 2009. Therefore, we forecasted the value of the average income of the two categories performing a linear regression. The results, available in a new dataframe, were later on bounded with the orginal one, in order to have the entire time series. 

*After that, we used the once complete data on inequality to create a simplified inequality indicator. This consists on divididing the top 10% avereage income of singaporeans by the bottom 90% average income. Consequenltly, the coefficient of the division  shows how many times higher the top 10% average singaporean's income is in comparison with the bottom 90% income of the bottom population.  

* For the number of cars we simply divided them into the categories provided in the data original set: cars, buses, etc. Originnaly they were in one column so we separate them in several ones to have the year as a unique identifier. Afterwards, we merged all the individual data frames in the new data frame shape.  

* For the utilization of transport public we only added the value of 0 to the LRT category from 1995 to 1998, This allowed us to complete the data set. 

#Descriptive and inferential statistics

## Creating a summary table

The table below shows the basic decriptive statistics for our variable included in the complete datasets. 
```{r, echo = FALSE, message=FALSE, warning = FALSE, results = "asis"}

# Creating nicer labels
labels1 <- c("gdp per capita", "inequality", "top 10% income", "bottom 90% income", "number of cars", "number of rental cars", "number of taxis", "number of buses", "number of motorbikes", "other vehicles", "bus usage", "mrt usage", "lrt usage" )
# Creating the table
stargazer(data.final, type = "latex",
         title = "General data summary", lable = "labels1",
        digits = 1, out = "General data summary.tex")
```

##Trends in gdp per capita, bottom 90% and top 10% average income
The following graph shows the trend of the three explanatory variables, throughout the period of time taken into consideration in Singapore. As we can see, 

```{r, echo = FALSE, message=FALSE, warning = FALSE}
# Creating first figure: trend in gdp, bottom 90% and top 10% average income
fig.1 <- ggplot2::ggplot(data.final, aes(x=data.final$date)) + 
  geom_line(aes(y=data.final$top, color="top")) + 
  geom_line(aes(y=data.final$bottom, color="bottom")) + 
  geom_line(aes(y=data.final$gdp.per.capita, color="gdp.per.capita")) + 
  theme_light()

#changing the name of the variables in the y axis
fig.1<-fig.1 + labs(x = "years", y = NULL)

#changing the position of the legend 
fig.1 <- fig.1 + theme(legend.position="bottom")

#changing the name of title and labels of legend
fig.1 <- fig.1 + scale_colour_discrete(name= NULL,
                         breaks=c("bottom", "gdp.per.capita", "top"),
                         labels=c("Bottom 90% average income", "Gdp per capita", "Top 10% average income"))

#changing the colour of the background of the plot
fig.1 <- fig.1 + theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.25, linetype = "solid",
                                   colour = "black"), 
          axis.line.y = element_line(size = 0.25, linetype = "solid",
                                   colour = "black"))
#giving a title to the graph
fig.1 <- fig.1 + ggtitle("Figure 1 - Gdp per capita, top 10% and bottom 90% average income\n
in Singapore measured in national currency at current prices (1995-2014)") + theme(plot.title = element_text(lineheight=0.5, size = 10))

fig.1
```

```{r, echo = FALSE, message=FALSE, warning = FALSE}
#creating the second figure: trend in inequality in number of times. 
fig.2 <- ggplot2::ggplot(data.final, aes(x=data.final$date, y=data.final$inequality)) + geom_line(aes(group=1), colour="#3399FF") + theme_classic()

#changing the name of the variables in the y axis
fig.2 <-fig.2 + labs(x = "years", y = "inequality")

#changing the colour of the background of the plot
fig.2 <- fig.2 + theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.25, linetype = "solid",
                                   colour = "black"), 
          axis.line.y = element_line(size = 0.25, linetype = "solid",
                                   colour = "black"))
                                   
#giving a title to the graph
fig.2 <- fig.2 + ggtitle("Figure 2 - Inequality in Singapore measured by the difference between \n the top 10% and bottom 90% average income in number of times (1995-2014)") + theme(plot.title = element_text(lineheight=1, size = 10))

fig.2
```

``````{r, echo = FALSE, message=FALSE, warning = FALSE}
#creating the third figure: trend in number of vehicles      
fig.3 <- ggplot2::ggplot(data.final, aes(x=data.final$date)) + 
  geom_line(aes(y=data.final$cars, color="cars")) + 
  geom_line(aes(y=data.final$rentalcars, color="rental cars")) + 
  geom_line(aes(y=data.final$taxis, color="taxis")) +
  geom_line(aes(y=data.final$buses, color="buses")) +
  geom_line(aes(y=data.final$motorbikes, color="motorbikes")) + 
  theme_classic()

#changing the name of the variables in the y axis
fig.3 <- fig.3 + labs(x = "years", y = NULL)

#changing the position of the legend 
fig.3 <- fig.3 + theme(legend.position="bottom")

#changing the name of title and labels of legend
fig.3 <- fig.3 + scale_colour_discrete(name= NULL,
                                       breaks=c("buses", "cars", "motorbikes", "rental cars", "taxis"),
                                       labels=c("buses", "cars", "motorbikes", "rental cars", "taxis"))

#changing the colour of the background of the plot
fig.3 <- fig.3 + theme(panel.border = element_blank(),
                       panel.grid.major = element_blank(),
                       panel.grid.minor = element_blank(),
                       axis.line.x = element_line(size = 0.25, linetype = "solid",
                                                  colour = "black"), 
                       axis.line.y = element_line(size = 0.25, linetype = "solid",
                                                  colour = "black"))
#giving a title to the graph 
fig.3 <- fig.3 + ggtitle("Figure 3 - Number of public and private vehicles in Singapore (1995-2014)") + theme(plot.title = element_text(lineheight=1, size = 10))

#changing the scale of the y axis
fig.3 <- fig.3 + scale_y_continuous(name= NULL, labels = comma)

fig.3                                  
```

```{r, echo = FALSE, message=FALSE, warning = FALSE}
#creating the fourth figure: trend in vehicles utilization           
fig.4 <- ggplot2::ggplot(data.final, aes(x=data.final$date)) + 
  geom_line(aes(y=data.final$bus.u, color="buses")) + 
  geom_line(aes(y=data.final$mrt.u, color="MRT")) + 
  geom_line(aes(y=data.final$lrt.u, color="LRT")) + theme_light()

#changing the name of the variables in the y axis
fig.4 <-fig.4 + labs(x = "years", y = NULL)

#changing the colour of the background of the plot
fig.4 <- fig.4 + theme(panel.border = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.line.x = element_line(size = 0.25, linetype = "solid",
                                   colour = "black"), 
          axis.line.y = element_line(size = 0.25, linetype = "solid",
                                   colour = "black"))
##changing the position of the legend 
fig.4 <- fig.4 + theme(legend.position="bottom")

#changing the name of title and labels of legend
fig.4 <- fig.4 + scale_colour_discrete(name= NULL,
                         breaks=c("buses", "LRT", "MRT"),
                         labels=c("buses", "MRT", "LRT"))
fig.4 <- fig.4 + ggtitle("Figure 4 - Average daily passengers on public transports in Singapore \n in thousands (1995-2014)") + theme(plot.title = element_text(lineheight=1, size = 10))
fig.4
```

## Multiple regression analysis

```{r, echo = FALSE, message=FALSE, warning = FALSE}
# Creating the firts model
m1 <- lm(cars ~ gdp.per.capita + inequality + bus.u + mrt.u + lrt.u, data = data.final)
m2 <- lm(bus.u ~ gdp.per.capita + inequality, data = data.final)
m3 <- lm(mrt.u ~ gdp.per.capita + inequality, data = data.final)
m4 <- lm(buses ~ gdp.per.capita + inequality + bus.u + mrt.u + lrt.u,data = data.final)
```

```{r, echo = FALSE, message=FALSE, warning = FALSE}
# Creating a table of the models with stargazer
stargazer::stargazer(m1, m2, m3, m4,  
    title = 'Multiple regeression models',
    digits = 2, type = 'latex', header =  FALSE)
```


#References

\pagebreak
#Annex
## Annex 1: Final data set containing all the variables and observations
```{r, echo = FALSE, message=FALSE, warning = FALSE}
pander::pander(data.final, split.cells = 70, split.table = Inf, style = "grid")
```
